DRAFT: Direct-style process creation on Linux

### 2020-05-15 research/paper/paper.tex
paper

Our direct-style process creation implementation uses \texttt{clone} to create a child process in an embryonic state
which can then be manipulated from user code using syscalls with \texttt{rsyscall}.
A direct-style \texttt{clone} has all the advantages of direct-style process creation
over Linux's fork-style \texttt{fork} syscall and spawn-style \texttt{posix\_spawn} libc function.
Direct-style \texttt{clone} can be used correctly and performantly from any process.
Cross-process syscalls made with rsyscall individually report errors, as normal, and can be easily branched on.
Arbitrary processes involving complex namespaces and initialization can be created
using techniques that are impractical with \texttt{fork}.
The performance cost from high-level code varies, but the underlying primitives are substantially faster.

@@

This paper is organized in several sections.
In section \ref{background}, we provide background on fork-style, spawn-style, and direct-style process creation,
and consider their advantages and disadvantages in more depth.
In section \ref{overview}, we discuss the design of the API,
including our usage of \texttt{rsyscall}.
In section \ref{applications}, we show examples of usage of direct-style process creation in Python on Linux,
including use of process features which are difficult to use on Linux with \texttt{fork} or \texttt{posix\_spawn}.
In section \ref{implementation}, we explore our implementation in depth,
and discuss several difficult process-creation related details of \texttt{rsyscall}'s implementation.
In section \ref{evaluation}, we evaluate, somehow (TODO).
In section \ref{related_work},
we examine other work which also touches on the topics covered by this paper.
In section \ref{future_work},
we list several directions for future work related to direct-style process creation and to \texttt{rsyscall}.
In section \ref{conclusions}, we briefly and optimistically conclude.

\section{Background}\label{background}
Processes are a widespread feature of operating systems,
with substantial variation in characteristics between systems.
One of the areas of variation is in process creation mechanisms.

On most systems today,
the interfaces for process creation
can be divided into "fork-style" and "spawn-style".
A few systems use a third style, which we refer to as "direct-style".
\subsection{Fork-style}
By far the most well-known fork-style interface is Unix's fork.\cite{forkhist}
With a fork-style interface,
a process, called the parent process, calls \texttt{fork},
and a copy of that process is made,
called the child process,
differing only in a few attributes such as process id.
The same program continues executing in both processes,
with a different return value from fork to allow it to distinguish whether it is running in the parent or the child.
Typically, the program will case on the return value
and, if running in the child,
call various system calls to modify the state of the new process.

@@

In direct-style, a process is created by its parent in a non-running state,
then supplied with various resources,
and then started running once it is fully set up.
In operating systems with true direct-style process creation,
the syscalls that can mutate a process
take explicit arguments to indicate the process on which they should operate.
In this way, the same syscalls that can mutate a process while it is running,
are called by the parent process to mutate the process while it is being set up.

@@

The familiar and mundane nature of the API is important not just for the learning curve,
but also for expressivity.
Since the API is simply normal Linux system calls,
direct-style clone is at least as expressive as fork and syscalls;
and, in fact, as we'll see in section \ref{applications},
is significantly more expressive.
\section{Applications}\label{applications}
In this section,
we'll discuss several more advanced uses of processes,
using more sophisticated Linux features,
and accompany them with Python examples using direct-style process creation.

@@

\lstinputlisting[float=*,language=Python,label={fuse},caption={Shared file descriptor tables}]
{examples/fuse/direct.py}
In listing \ref{fuse},
we first create a child process, \verb|ns_child|, in a new mount namespace with \verb|CLONE_NEWNS|,
and for the first time, also pass \verb|CLONE_FILES|,
which causes the file descriptor table to be shared between the parent process and the child process.
We create a grandchild off of \verb|ns_child|,
and use it to exec a FUSE filesystem,
which will appear only in the mount namespace of \verb|ns_child| and its descendents.
Then we can open FUSE files in \verb|ns_child|
and use those files in the parent process,
through the shared file descriptor table.

@@

Alternatively, some way to wait for futex wakeups through a file descriptor could be added,
so we can use file-descriptor-monitoring syscalls to wait for the \texttt{ctid} futex;
such a feature used to exist in the form of \verb|FUTEX_FD|,
but was removed from Linux long ago due to race conditions in its design.
\subsection{Managing file descriptor tables}\label{fdtables}
As demonstrated in section \ref{shared_fd_table},
the \verb|CLONE_FILES| flag can be passed to \texttt{clone},
causing the file descriptor table to be shared between the parent process and child process.
The same file descriptors are open in both processes at the same numbers,
and if new file descriptors are opened in either process,
they are also visible in the other process.
This is simple to model, and convenient for many purposes.
The file descriptor table is unshared once the child calls \texttt{execve}.

>>>>

Our direct-style process creation implementation uses \texttt{clone} to create a child process in an embryonic state
which can then be manipulated from user code using syscalls with \texttt{rsyscall}.
A direct-style \texttt{clone} has all the advantages of direct-style process creation
over Linux's fork-style \texttt{fork} syscall and spawn-style \texttt{posix\_spawn} libc function.
\begin{itemize}
\item
Direct-style \texttt{clone} can be used correctly and performantly from any process.
\item
Cross-process syscalls made with rsyscall individually report errors, as normal, and can be easily branched on.
\item
Arbitrary processes involving complex namespaces and initialization can be created
using techniques that are impractical with \texttt{fork}.
\item
The performance cost from high-level code varies, but the underlying primitives are substantially faster.
\end{itemize}

@@

This paper is organized in several sections.
In section \ref{background}, we provide background on fork-style, spawn-style, and direct-style process creation,
and consider their advantages and disadvantages in more depth.
In section \ref{overview}, we discuss the design of the API,
including our usage of \texttt{rsyscall}.
In section \ref{applications}, we show examples of usage of direct-style process creation in Python on Linux,
including use of process features which are difficult to use on Linux with \texttt{fork} or \texttt{posix\_spawn}.
In section \ref{implementation}, we explore our implementation in depth,
and discuss several difficult process-creation related details of \texttt{rsyscall}'s implementation.
In section \ref{evaluation}, we evaluate, somehow (TODO).
In section \ref{related_work},
we examine other work which also touches on the topics covered by this paper.
In section \ref{future_work},
we list several directions for future work related to direct-style process creation and to \texttt{rsyscall}.
% TODO
In section \ref{conclusions}, we briefly and optimistically conclude.

\section{Background}\label{background}
On most systems today,
the interfaces for process creation
can be divided into "fork-style" and "spawn-style".
A few systems use a third style, which we refer to as "direct-style".
\subsection{Fork-style}
By far the most well-known fork-style interface is Unix's fork.\cite{forkhist}
With a fork-style interface,
a process, called the parent process, calls \texttt{fork},
and a copy of that process is made,
called the child process,
differing only in a few attributes such as process id.
The same program continues executing in both processes,
with a different return value from fork to allow it to distinguish whether it is running in the parent or the child.
Typically, the program will case on the return value
and, if running in the child,
call various system calls to modify the state of the new process.

@@

In direct-style, a process is created by its parent in a non-running state,
then supplied with various resources,
and then started running once it is fully set up.
In operating systems with true direct-style process creation,
the syscalls that can mutate a process
take explicit arguments to indicate the process on which they should operate.
In this way, the same syscalls that can mutate a process while it is running
are called by the parent process to mutate the process while it is being set up.

@@

The familiar and mundane nature of the API is important not just for the learning curve,
but also for expressivity.
Since the API is simply normal Linux system calls,
direct-style clone is at least as expressive as fork and syscalls;
and, in fact, as we'll see in section \ref{applications},
is significantly more expressive.
\section{Applications}\label{applications}
In this section,
we'll discuss several more advanced uses of processes
using more sophisticated Linux features,
and accompany them with Python examples using direct-style process creation.

@@

\lstinputlisting[float=*,language=Python,label={fuse},caption={Shared file descriptor tables}]
{examples/fuse/direct.py}
In listing \ref{fuse},
we first create a child process \verb|ns_child| in a new mount namespace with \verb|CLONE_NEWNS|,
and, for the first time, also pass \verb|CLONE_FILES|,
which causes the file descriptor table to be shared between the parent process and the child process.
We create a grandchild off of \verb|ns_child|
and use it to exec a FUSE filesystem,
which will appear only in the mount namespace of \verb|ns_child| and its descendents.
Then we can open FUSE files in \verb|ns_child|
and use those files in the parent process,
through the shared file descriptor table.

@@

Alternatively, some way to wait for futex wakeups through a file descriptor could be added,
so we can use file-descriptor-monitoring syscalls to wait for the \texttt{ctid} futex;
such a feature used to exist in the form of \verb|FUTEX_FD|,
but was removed from Linux long ago due to race conditions in its design.
\subsection{Managing file descriptor tables}\label{fdtables}
As demonstrated in section \ref{shared_fd_table},
the \verb|CLONE_FILES| flag can be passed to \texttt{clone},
causing the file descriptor table to be shared between the parent process and child process.
The same file descriptors are open in both processes at the same numbers,
and if new file descriptors are opened in either process,
they are also visible in the other process.
This is simple to model and convenient for many purposes.
The file descriptor table is unshared once the child calls \texttt{execve}.
